---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# coalRate

<!-- badges: start -->
<!-- badges: end -->

The goal of coalRate is to provide easily accessible methods for estimating the 
growth rate of clones using time-based phylogenetic trees as input. This package 
provides the internal lengths and maximum likelihood methods from our recent 
preprint [Estimating single cell clonal dynamics in human blood using coalescent 
theory](https://www.biorxiv.org)

## Installation

You can install the development version of coalRate from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bdj34/coalRate")
```

## Example

### Simulated data

This is a basic example which shows you how to simulate a tree and calculate the 
growth rate using two different methods:

```{r sim example}
library(coalRate)

# Generate a sampled tree with 100 tips from a 20 year birth-death process with birth rate a=1 and death rate b=0.5
tree <- simTree(a=1, b=0.5, cloneAge=20, n=100)

# Plot the tree (see ggtree for more advanced plotting)
plot(tree)

# Estimate the growth rate r=a-b=0.5 using maximum likelihood
maxLike.df <- maxLikelihood(tree)
print(paste0("Max. likelihood estimate = ", maxLike.df$estimate))

# Estimate the growth rate r=a-b=0.5 using internal lengths
intLengths.df <- internalLengths(tree)
print(paste0("Internal lengths estimate = ", intLengths.df$estimate))
```

In our [paper](https://www.biorxiv.org), we use simulated trees to test our 
growth rate estimates.

### Real data

You can apply these methods to real data as well. Here, we apply our maximum 
likelihood and lengths estimates to a real data clone from [Williams et al.](https://pubmed.ncbi.nlm.nih.gov/35058638/)

```{r realData example}
library(coalRate)

# Load and plot the data
PD9478 <- coalRate::realCloneData[["PD9478_1"]]
plot(PD9478)
```

Let's now use the expansions data.frame to subset the tree using the ape function keep.tip()
```{r get subclone}
# Select the subclone identified in the expansions data.frame
cloneTree <- ape::keep.tip(PD9478, PD9478$expansions$Tips[PD9478$expansions$Expansion == "clone1"])
plot(cloneTree)
```

We can now apply our methods to the clone tree
```{r apply methods}
# Get maximum likelihood and internal lengths estimates
print(maxLikelihood(cloneTree)$estimate)
print(internalLengths(cloneTree)$estimate)
```


Our package comes with 42 clones annotated from four distinct publications, 
which are the ones we use in our analysis. Note that there are three clones
profiled at two different timepoints, meaning there are 39 unique clones.
The papers which generate this data are:

[Williams et al. 2022](https://pubmed.ncbi.nlm.nih.gov/35058638/)

[Mitchell et al. 2022](https://pubmed.ncbi.nlm.nih.gov/35650442/)

[Fabre et al. 2022](https://pubmed.ncbi.nlm.nih.gov/35650444/)

[Van Egeren et al. 2021](https://pubmed.ncbi.nlm.nih.gov/33621486/)




