---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# coalRate

<!-- badges: start -->
<!-- badges: end -->

The goal of coalRate is to provide easily accessible methods for estimating the 
growth rate of clones using time-based phylogenetic trees as input. This package 
provides the internal lengths and maximum likelihood methods from our recent 
preprint [Estimating single cell clonal dynamics in human blood using coalescent 
theory](https://www.biorxiv.org)

## Installation

You can install the development version of coalRate from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bdj34/coalRate")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(coalRate)

# Generate a sampled tree with 100 tips from a 20 year birth-death process with birth rate a=1 and death rate b=0.5
tree <- simTree(a=1, b=0.5, cloneAge=20, n=100)

# Estimate the growth rate r=a-b=0.5 using maximum likelihood
result.df <- maxLikelihood(tree)
print(result.df$estimate)
```

You can apply these methods to real data as well. Here, we apply our maximum 
likelihood and lengths estimates to a real data clone from [Williams et al.](https://pubmed.ncbi.nlm.nih.gov/35058638/)

```{r cars}
library(coalRate)

```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
